Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-11-03 pituba/article/article_spe.tex
corrections

In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between the volume of dissolved gas and the volume of oil both measured in surface condition:

@@

Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at surface condition and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil phases measured in reservoir condition.

@@

It is now possible to obtain the so-called generalized material balance equation by substituting the equations for phase saturation, Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5}, into Eq.~\eqref{eq: S15}:

@@

It can be shown that neglecting the variation of density with depth does not introduce  important error. As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$.

@@

For investigating the order of magnitude of Eqs.~\eqref{eq: gradodz} and \eqref{eq: gradgdz}, we substitute into them the values of $\rho_o = 10^3\ kg/m^3$ and $c_o = 10^{-9}\ Pa^{-1}$ for the oil phase, and $\mathcal{M}=19\ kg/kmol$ and $T=400\ K$ for the gas phase: 
\begin{equation}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 1\times10^{-5}\ m^{-1} \, ,
\end{equation}

@@

\section{Results}
For illustrating the numerical procedure, the material balance equation was solved for a fictitious black-oil reservoir initially producing in a pure depletion drive mechanism and later subjected to water injection. Fig.~\ref{fig: matbal_p} shows in the solid line curve the severe depletion during initial phases of production. In the dashed curve, Fig.~\ref{fig: matbal_p} shows the bubble-point pressure, indicating how far the saturated reservoir is from undersaturation and allowing for planning of pressure maintenance projects.

@@

\section*{Nomenclature}
\begin{tabular}{ll}
$B_g$ & gas formation volume factor\\
$B_o$ & oil formation volume factor\\
$B_w$ & water formation volume factor\\
$R_s$ & solubility ratio of gas in oil phase\\
$S_g$ & gas saturation \\
$S_o$ & oil saturation \\
$S_w$ & water saturation \\
$G_p$ & produced volume of gas \\
$m$ & ratio of original free gas volume \\
& to original oil volume in reservoir condition \\
$N$ & original volume of oil in place \\
$N_p$ & produced volume of oil \\
$P_b$ & bubble-point pressure \\
$V_p$ & porous volume \\
$W_i$ & injected volume of water \\
$W_p$ & produced volume of water
\end{tabular}
\section*{References}
\bibliographystyle{elsarticle-harv}
\bibliography{refs}
\cite{*}

>>>>

In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between the volume of dissolved gas and the volume of oil both measured at surface condition:

@@

Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at surface condition and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil phases measured at reservoir condition.

@@

It is now possible to obtain the generalized material balance equation by substituting the equations for phase saturation, Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5}, into Eq.~\eqref{eq: S15}:

@@

It can be shown that neglecting the variation of density with depth does not introduce  important errors. As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$.

@@

To investigate the order of magnitude of Eqs.~\eqref{eq: gradodz} and \eqref{eq: gradgdz}, we substitute into them the values of $\rho_o = 10^3\ kg/m^3$ and $c_o = 10^{-9}\ Pa^{-1}$ for the oil phase, and $\mathcal{M}=19\ kg/kmol$ and $T=400\ K$ for the gas phase: 
\begin{equation}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 1\times10^{-5}\ m^{-1} \, ,
\end{equation}

@@

\section{Results}
To illustrate the numerical procedure, the material balance equation was solved for a fictitious black-oil reservoir initially producing in a pure depletion drive mechanism and later subjected to water injection. Fig.~\ref{fig: matbal_p} shows in the solid line curve the severe depletion during initial phases of production. In the dashed curve, Fig.~\ref{fig: matbal_p} shows the bubble-point pressure, indicating how far the saturated reservoir is from undersaturation and allowing for planning of pressure maintenance projects.

@@

\section*{Nomenclature}
\begin{tabular}{ll}
$B_g$ & gas formation volume factor\\
$B_o$ & oil formation volume factor\\
$B_w$ & water formation volume factor\\
$R_s$ & solubility ratio of gas in oil phase\\
$S_g$ & gas saturation \\
$S_o$ & oil saturation \\
$S_w$ & water saturation \\
$G_p$ & produced volume of gas \\
$m$ & ratio of original free gas volume \\
& to original oil volume at reservoir condition \\
$N$ & original volume of oil in place \\
$N_p$ & produced volume of oil \\
$P_b$ & bubble-point pressure \\
$V_p$ & porous volume \\
$W_i$ & injected volume of water \\
$W_p$ & produced volume of water
\end{tabular}
\section*{References}
\bibliographystyle{elsarticle-harv}
\bibliography{refs}
\cite{*}
